---
# UNL thesis fields
title: "STATISTICAL TESTING OF ACTIVITY CLIFFS"
author: "Sarah Josephine Aurit"
month: "December"
year: "2023"
location: "Lincoln, Nebraska"
major: "Statistics"
adviser: "Souparno Ghosh"
abstract: |
  Here is my abstract. *(350 word limit)*
acknowledgments: |
  Thank you to all my people!
dedication: |
  Dedicated to...
# End of UNL thesis fields
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    pandoc_args: --top-level-division=chapter
    keep_tex: yes
    latex_engine: xelatex
    template: template.tex
  huskydown::thesis_gitbook: 
    style: style.css
#  huskydown::thesis_word: default
#  huskydown::thesis_epub: default
bibliography: bib/thesis.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: bib/apa.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
#header-includes:
#- \usepackage{tikz}
---

```{r setup, include = F}
options(width = 60)
knitr::opts_chunk$set(
  echo = F, 
  eval = T, 
  messages = F, 
  warnings = F,
  fig.width = 6, 
  fig.height = 4,  
  fig.align = 'center',
  out.width = "\\linewidth", 
  dpi = 300, 
  tidy = T, tidy.opts=list(width.cutoff=45),
  fig.pos = "tbp",
  out.extra = "",
  cache = FALSE
)
```

```{r include_packages, include = FALSE}
# This chunk ensures that the huskydown package is
# installed and loaded. This huskydown package includes
# the template files for the thesis.
if(!require(devtools))
  install.packages("devtools", 
                   repos = "http://cran.rstudio.com")
if(!require(huskydown))
  devtools::install_github(
    "benmarwick/huskydown"
  )
library(huskydown)
library(tidyverse)
library(knitr)
library(ggplot2)
```

# Literature Review

## Motivation and Background

The journey of discovering effective cancer drugs can be long and expensive. The elements of time and cost weigh against the race to improve survival and the quality of life for cancer patients. Researchers typically start testing drug compounds on cancer cell lines within a laboratory setting first before testing on animals or humans. To narrow the search field, drugs are sometimes pursued that are similar to ones previously found to be successful in cancer treatment. That said, it is possible for the new drug to be much less effective, which is defined as an activity cliff (Cruz-Monteagudo et al., 2014).

According to HU et al. (2017), compounds that are structurally similar yet have large potency differences are identified as producing activity cliffs upon an activity landscape. Additionally, an activity cliff is potentially produced within the context of the following four key aspects: 1) we are considering and comparing a pair of compounds, 2) both compounds are active against the same target, 3) the two compounds are similar in chemical structure, and 4) there is a difference in potency.

Focusing on the aspect of compound similarity via cheminformatics, formal investigation is completed by looking at similarity measurements. There are multiple approaches regarding the quantification of similarity between two compounds. One popular selection is the Tanimoto coefficient.

The Tanimoto coefficient, a measure of similarity, is determined by the quantification of molecular attributes that are similar between two molecular compounds. Essentially, it is a ratio of the number of similar entities divided by the total number of molecular attributes. "Why is Tanimoto index an appropriate choice for fingerprint-based similarity calculations?"

Analogously, matched molecular pairs (MMP) analysis encompasses another measurement of similarity. Essentially, two molecules differ by one "well-defined, structural" chemical transformation. https://pubs-acs-org.libproxy.unl.edu/doi/pdf/10.1021/jm200452d. MMPs were initially used as a special case of a quantitative structure-activity relationship (QSAR), which relates molecular compounds to the response of potency or other biochemical properties.

QSAR methodology can be essential in determining the relationships between chemical structures and biochemical activities, and it is typical that the methodology incorporates supervised machine learning techniques. Compounds are first represented by a numerical entity such as a two-dimensional structural formula that promulgates atom connectivity via chemical bonds. The structural formula then is related to bioactivities. It is not uncommon for there to be violations of QSAR methodology given the presence of an activity cliff (AC). Reference: QSAR modeling where have you been where are you going

The Genomics of Drug Sensitivity in Cancer (GDSC) database was formed based on a collaborative effort of the UK's Wellcome Sanger Institute's Cancer Genome Project and Massachusetts' General Hospital's Center for Molecular Therapeutics. It is a repository and compilation of anti-cancer drug response information linked to cancer cell lines of multiple primary cancer sites. The response is quantified by \textit{$\mathrm{IC_50}$}, a metric that reveals how much substance or dose is necessary to inhibit a biological component by 50%. Also, AUC values are calculated from dose-response curves attributed to nine dose levels where drug response is measured 72 hours post-administration of drug to cancer cell lines. Disregarding missing data, approximately 900 drug characteristics were identified within this data source.

Look at data

DFTs








